A Classical Density-Functional Theory for Describing Water Interfaces

### 2012-06-05 papers/water-SAFT/paper.tex
Changed coloring/size on single rod density plot.  Made edits to paper. Tested everything on node07.

As common as water is, its large dielectric constant, small
molecular size, strong hydrogen-bonding network and anomolous
thermodynamic properties make ``the universal solvent'' an interesting
subject of study.

@@

A number of classical density functionals have been developed for
water~\cite{ding1987, Yang1992, gloor2002saft,gloor2004accurate,
  gloor2007prediction, Jaqaman2004, clark2006developing,
  lischner2010classical, fu2005vapor-liquid-dft,felipe2001examination}
, capturing some of the qualitative behavior of water.
However, many of these are less sophisticated than those describing
the hard-sphere fluid, and a quantitatively accurate functional
describing water---which reproduces the temperature dependence of the
macroscopic surface tension, its decrease as curvature increases, and
its quantitative value---has not yet been published.  For instance,
the functional of Lischner \emph{et al}\cite{lischner2010classical}
treats the surface tension correctly, but can only be used at room
temperature.  A number of classical density functionals have recently
been produced that are based on Statistical Associating Fluid Theory
(SAFT)\cite{gloor2002saft, clark2006developing, gloor2007prediction,
  gloor2004accurate, gross2009density, kahl2008modified, felipe2001examination}.
These functionals are based on a perturbative thermodynamic
expansion, and give a reasonable prediction for the temperature
dependence of the surface tension.  However, they also overestimate
the surface tension by a factor of about 1.5, and \fixme{most?} are unsuited
for studying systems that have density variations on a molecular
length scale due to the use of a local density approximation.

@@

The SAFT Helmholtz free energy is composed of five terms:
\begin{equation} \label{eq:SAFT-free-energy}
  F[n] = F_\textit{id} + F_\textit{hs} + F_\textit{disp} +
  F_\textit{assoc} + F_\textit{chain},
\end{equation}
where the first three terms encompass the \emph{monomer} contribution
to the free energy, the fourth is the \emph{association} free energy,
describing hydrogen bonds, and the final term is the chain formation
energy of for fluids that are chain polymers.  The final term is not
relevant for water.  While a number of SAFT formulations have been
published, we will focus on SAFT-VR\cite{gil-villegas-1997-SAFT-VR},
which has been used by Clark \emph{et al} to construct an optimal SAFT
model for water\cite{clark2006developing}.

@@

SAFT has also been used to construct classical density functionals,
which have been applied to study the surface tension as a function of
temperature\cite{whotocite}.  Such functionals have qualitatively
predicted the dependence of surface tension on temperature, but have
in general provided poor quantitative agreement with experiment.
Moreover, the vast majority of SAFT-based classical density
functionals have made local-density approximations that make them
unsuitable for application to systems that obey the contact-value
theorem. \cite{felipe2001examination, gloor2002saft,
  gloor2004accurate, clark2006developing, gloor2007prediction,
  kahl2008modified, gross2009density}.\

 However, they also overestimate
the surface tension by a factor of about 1.5, and \fixme{most?} are unsuited
for studying systems that have density variations on a molecular
length scale due to the use of a local density approximation.

Functionals constructed using a local density approximation
  will always fail to satisfy the contact-value theorem, and will
  therefore incorrectly model small hard solutes.
The contact-value theorem states that the pressure of any fluid at a
hard wall interface is proportional to the contact density of the
fluid\cite{henderson1979exact}, according to the following relation:
\begin{equation}\label{eq:contact}
  p = n_ck_BT
\end{equation}
where $n_c$ is the contact density, $k_B$ is the Boltzmann constant
and $T$ is the temperature of
the fluid. This leads to the property that the free energy of
solvation of a small hard solute is proportional to the
solvent-excluded volume:
\begin{equation}\label{contactvaluethm}
  F = n k_BT V
\end{equation}
\fixme{add this here? :} The contact-value theorem is satisfied for
functionals for with the only purely local term is the free energy of
an ideal gas.

@@

This is consistent with the calculation of the surface tension in
Figure~\ref{fig:surface-tension} (72~mN/m) and the force per length
found from the slopes of the curves in Figure
~\ref{fig:rods-energy-vs-distance} (144~mN/m)  which
compares the free energy per length for rods of several different
radii. This agreement indicates that even though the rods are nanoscale, the
water molecules will ``see'' them classically.  For large $d$, the rods 
simply act like two separate single rods in water and the energy will
reach an asymptote determined by the effective surface tension (shown in 
Figure~\ref{fig:energy-vs-diameter}). In Figure
~\ref{fig:rods-energy-vs-distance}, we see that for small $d$ all the rods 
have the same behavior with a positive linear relationship
between free energy per length and separation during the initial state 
(where the rods have vapor between them). This indicates that the force 
per length on the rods is constant, and surprisingly independent of separation.

>>>>

As common as water is, its large dielectric constant, small
molecular size, strong hydrogen-bonding network and anomolous
thermodynamic properties make ``the universal solvent'' a unique
and interesting material.

@@

A number of classical density functionals have been developed for
water~\cite{ding1987, Yang1992, gloor2002saft,gloor2004accurate,
  gloor2007prediction, Jaqaman2004, clark2006developing,
  lischner2010classical, fu2005vapor-liquid-dft,felipe2001examination}
, capturing some of the qualitative behavior of water.
However, many of these are less sophisticated than those describing
the hard-sphere fluid, and a quantitatively accurate functional
describing water---which reproduces the temperature dependence of the
macroscopic surface tension, its decrease as curvature increases, and
its quantitative value---has not yet been published.  For instance,
the functional of Lischner \emph{et al}\cite{lischner2010classical}
treats the surface tension correctly, but can only be used at room
temperature.  A number of classical density functionals have recently
been produced that are based on Statistical Associating Fluid Theory
(SAFT)\cite{gloor2002saft, clark2006developing, gloor2007prediction,
  gloor2004accurate, gross2009density, kahl2008modified, felipe2001examination}.
These functionals are based on a perturbative thermodynamic
expansion, and give a reasonable prediction for the temperature
dependence of the surface tension.

@@

The SAFT Helmholtz free energy is composed of five terms:
\begin{equation} \label{eq:SAFT-free-energy}
  F[n] = F_\textit{id} + F_\textit{hs} + F_\textit{disp} +
  F_\textit{assoc} + F_\textit{chain},
\end{equation}
where the first three terms encompass the \emph{monomer} contribution
to the free energy, the fourth is the \emph{association} free energy,
describing hydrogen bonds, and the final term is the chain formation
energy of for fluids that are chain polymers.  The final term is not
relevant for water.  While a number of SAFT formulations have been
published, we will focus on SAFT-VR\cite{gil-villegas-1997-SAFT-VR},
which has been used by Clark \emph{et al} to construct an optimal SAFT
model for water\cite{clark2006developing}. A majority of the empirical
parameters used in our functional are taken from the Clark\emph{et al}
SAFT functional which contains five empirical parameters and give the
correct pressure at liquid-vapor coexistince curve shown in Figure
\ref{fig:pressure-with-isotherms}.

@@

SAFT has also been used to construct classical density functionals,
which have been applied to study the surface tension as a function of
temperature\cite{whotocite}.  Such functionals have qualitatively
predicted the dependence of surface tension on temperature, however, they also overestimate
the surface tension by a factor of about 1.5, and \fixme{most?} are unsuited
for studying systems that have density variations on a molecular
length scale due to the use of a local density approximation.

Functionals constructed using a local density approximation
  will always fail to satisfy the contact-value theorem, and will
  therefore incorrectly model small hard solutes.
The contact-value theorem states that the pressure of any fluid at a
hard wall interface is proportional to the contact density of the
fluid\cite{henderson1979exact}, according to the following relation:
\begin{equation}\label{eq:contact}
  p = n_ck_BT
\end{equation}
where $n_c$ is the contact density, $k_B$ is the Boltzmann constant
and $T$ is the temperature of
the fluid. This leads to the property that the free energy of
solvation of a small hard solute is proportional to the
solvent-excluded volume:
\begin{equation}\label{contactvaluethm}
  F = n k_BT V
\end{equation}
The contact-value theorem is satisfied for functionals where the
only purely local term is the free energy of an ideal gas.

@@

This is consistent with the calculation of the surface tension in
Figure~\ref{fig:surface-tension} (72~mN/m) and the force per length
found from the slopes of the curves in Figure
~\ref{fig:rods-energy-vs-distance} (144~mN/m)  which
compares the free energy per length for rods of several different
radii. This agreement indicates that even though the rods are nanoscale, the
water molecules will ``see'' them classically.  For large $d$, the rods 
simply act like two separate single rods in water and the energy will
reach an asymptote determined by the effective surface tension (shown in 
Figure~\ref{fig:energy-vs-diameter}). In Figure
~\ref{fig:rods-energy-vs-distance}, we see that for small $d$ all the rods 
have the same behavior with a positive linear relationship
between free energy per length and separation during the initial state 
with vapor between them. This indicates that the force 
per length on the rods is constant, and surprisingly independent of separation.
